### 2013-03-25 doc/dissertation/document/chapters/mc_theory/mc_theory.tex
adding additional pngs to symmetric spectral analysis for communication

\chapter{An Analytic Performance Framework for Domain-Decomposed Monte Carlo}
\label{ch:analytic_framework}

%%---------------------------------------------------------------------------%%
\section{Leakage Fractions for Symmetric Systems}
\label{sec:symmetric_leakage_fractions}
With Wagner's work, we observed that domain overlap implemented with a
domain decomposed Monte Carlo setting reduces the amount of nearest
neighbor communication that must occur during transport.

To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
Carlo applications, however, domain decomposition has been identified
as a key principle in moving forward in high performance computing to
enable higher fidelity simulations
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because these nearest-neighbor
communication sequences, specifically, have poor algorithmic strong
scaling \cite{gropp_high-performance_2001}.

The purpose of this study is to provide a simple, analytic theory
based on the properties of the linear system that will allow for
estimates of the domain decomposed behavior of the adjoint
Neumann-Ulam method. When solving problems where the linear operator
is symmetric, a host of analytic theories exist based on the
eigenvalue spectrum of the operator that characterize their behavior
in the context of deterministic linear solvers. Using past work, these
theories are adapted to the domain decomposed adjoint Neumann-Ulam
method using the one-speed, two-dimensional neutron diffusion
equation. In this paper we describe the adjoint Neumann-Ulam Monte
Carlo method followed by a presentation of the model problem. Using
the linear system generated by the discretization of the model
problem, we use a spectral analysis to generate analytic relations for
the eigenvalues of the operator based on system parameters. Using the
eigenvalue spectra, we then build relationships to characterize the
transport of stochastic histories in a decomposed domain and the
fraction of histories that leak from a domain and will therefore have
to be communicated. Finally, we compare these analytic results to
numerical experiments conducted with the model problem and draw
conclusions looking towards future work.

\subsection{Model Problem}
\label{subsec:model_problem}
For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{chapters/mc_theory/stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

@@

\subsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

@@

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
preconditioning is sufficient to reduce the spectral radius of the
iteration matrix below unity and therefore ensure convergence of the
Neumann series. The preconditioner in this case is then $\ve{M} =
diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor, $\alpha$, is the
same for all rows in the operator and defined as the $\phi_{i,j}$
coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

The spectral radius of the iteration matrix is obtained by seeking its
largest eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. To find these maxium eigenvalue,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{chapters/mc_theory/diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}.
 We find that the maximum eigenvalue exists when $p=q=0$,
giving the following for the spectral radius of the Jacobi
preconditioned iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

\subsection{Neumann Series Convergence}
\label{subsec:neumann_convergence}
The adjoint Monte Carlo method is effectively an approximation to a
stationary method. Stationary methods for linear systems arise from
splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
\begin{equation}
  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
error at the $k^{th}$ iterate, the error after $k$ iterations is then:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
then have:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff, $W_c$, that determines
when low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

\subsubsection{Domain Leakage Approximations}
\label{subsubsec:domain_leak_approx}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, Siegel and colleagues assumed that the value of
$\Lambda$ was dependent on the total cross section of the system via
the Wigner rational approximation. Outlined more thoroughly by Hwang's
chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
rational approximation and the mean chord approximation as a means to
estimate the leakage fraction.

@@

\subsection{Numerical Experiments}
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4in,clip]{chapters/mc_theory/measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        section to scattering cross section ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

\subsubsection{Random Walk Length}
\label{subsubsec:walk_length}
With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source of strength 1
in the center of the domain. For each of the histories, the number of
transitions made was tallied to provide an effective value of $k$ for
each history. This value was then averaged over all histories to get a
measured value of $k$ for the particular operator. On the left,
Figure~\ref{fig:measured_length} presents these measurements as well
as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
the right, Figure~\ref{fig:measured_length} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.5in,clip]{chapters/mc_theory/measured_length.png}
    \end{center}
    \caption{\textbf{Measured and analytic random walk length as a
        function of the iteration matrix spectral radius.} \textit{The
        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
        \sn{1}{-8}. In the left plot, the red data was computed
        numerically by an adjoint Neumann-Ulam implementation while
        the black dashed data was generated by
        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
        error between the predicted and measured results is presented
        for each weight cutoff.}}
    \label{fig:measured_length}
  \end{spacing}
\end{figure}

>>>>

\chapter{An Analytic Performance Framework for Domain-Decomposed Monte Carlo}
\label{ch:analytic_framework}
With Wagner's work \citep{wagner_hybrid_2010}, we observed that domain
overlap implemented with a domain decomposed Monte Carlo setting
reduces the amount of nearest neighbor communication that must occur
during transport. To date, parallel Neumann-Ulam methods have been
limited to full domain replication with parallelism exploited through
individual histories \citep{alexandrov_efficient_1998} and in the
previous chapter we exploited Wagner's work to alleviate this. To
accomplish this, we recognized from the literature that stochastic
histories must be transported from domain to domain as the simulation
progresses and they transition to states that are not in the local
domain. Because we have chosen a domain decomposition strategy in a
parallel environment, this means that communication of these histories
must occur between compute nodes owning neighboring pieces of the
global domain as outlined in our algorithms. We wish to characterize
this communication not only because communication is in general
expensive, but also because these nearest-neighbor communication
sequences, specifically, have poor algorithmic strong scaling
\citep{gropp_high-performance_2001} in much the same way as a
matrix-vector multiply operation. Therefore, we desire a framework to
provide a simple, analytic theory based on the properties of the
linear system that will allow for estimates of the domain decomposed
behavior of the adjoint Neumann-Ulam method in terms of communication
costs.
%%---------------------------------------------------------------------------%%
\section{Leakage Fractions for Symmetric Systems}
\label{sec:symmetric_leakage_fractions}
When solving problems where the linear operator is symmetric, a host
of analytic theories exist based on the eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these theories are
adapted to the domain decomposed adjoint Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation. In this section
we use a one-speed two-dimensional neutron diffusion system as the
symmetric model problem. Using the linear system generated by the
discretization of the model problem, we use a spectral analysis to
generate analytic relations for the eigenvalues of the operator based
on system parameters. Using the eigenvalue spectra, we then build
relationships to characterize the transport of stochastic histories in
a decomposed domain and the fraction of histories that leak from a
domain and will therefore have to be communicated. Finally, we compare
these analytic results to numerical experiments conducted with the
model problem.

\subsection{Symmetric Model Problem}
\label{subsec:model_problem}
For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\citep{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\citep{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{chapters/mc_theory/stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

@@

\subsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\citep{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

@@

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
preconditioning as outlined in \S\ref{sec:stochastic_preconditioning}
is sufficient to reduce the spectral radius of the iteration matrix
below unity and therefore ensure convergence of the Neumann
series. Applying this preconditioner, we are then solving the
following diffusion system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
operator with each row scaled by the diagonal component. As we have
defined a homogeneous domain, the scaling factor, $\alpha$, is the
same for all rows in the operator and defined as the $\phi_{i,j}$
coefficient from Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we then have the following spectrum of
preconditioned eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

The spectral radius of the iteration matrix is obtained by seeking its
largest eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{multline}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}\Big]\:.
  \label{eq:iteration_stencil}
\end{multline}
Inserting the eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. To find these maxium eigenvalue,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{chapters/mc_theory/diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}
We find that the maximum eigenvalue exists when $p=q=0$, giving the
following for the spectral radius of the Jacobi preconditioned
iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

\subsection{Neumann Series Convergence}
\label{subsec:neumann_convergence}
As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
method is effectively an approximation to a stationary method. In the
adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\citep{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff, $W_c$, that determines
when low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

\subsubsection{Domain Leakage Approximations}
\label{subsubsec:domain_leak_approx}
In a domain decomposed situation, not all histories will remain within
the domain they started in and must instead be communicated. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations for full nuclear
reactor core Monte Carlo neutral particle simulations
\citep{siegel_analysis_2012}.  To quantify the number of particles
that leak out of the local domain they define a leakage fraction,
$\Lambda$, as:
\begin{equation}
  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, Siegel and colleagues assumed that the value of
$\Lambda$ was dependent on the total cross section of the system via
the Wigner rational approximation. Outlined more thoroughly by Hwang's
chapter in \citep{azmy_nuclear_2010}, we will use both the Wigner
rational approximation and the mean chord approximation as a means to
estimate the leakage fraction.

@@

\subsection{Numerical Experiments}
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=5in,clip]{chapters/mc_theory/measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        section to scattering cross section ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

\subsubsection{Random Walk Length}
\label{subsubsec:walk_length}
With the eigenvalue derivations verified, we can go about setting up
an experiment to measure the length of the random walks generated by
the adjoint Neumann-Ulam solver. To do this, we again use a $100
\times 100$ square grid with $h=0.1$ and the absorption cross varied
from 0 to 100 while the scattering cross section was fixed at
unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
\sn{1}{-8} were used with 10,000 histories generated by a point source
of strength 1 in the center of the domain. For each of the histories,
the number of transitions made was tallied to provide an effective
value of $k$ for each history. This value was then averaged over all
histories to get a measured value of $k$ for the particular
operator. Figure~\ref{fig:measured_length} presents these measurements
as well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as
a function of the iteration matrix spectral radius,
$\rho(\ve{H})$. FigureOn the right,
Figure~\ref{fig:measured_length_error} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length.png}
    \end{center}
    \caption{\textbf{Measured and analytic random walk length as a
        function of the iteration matrix spectral radius.} \textit{The
        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
        \sn{1}{-8}. The red data was computed numerically by an
        adjoint Neumann-Ulam implementation while the black dashed
        data was generated by Eq~(\ref{eq:analytic_k}).}}
    \label{fig:measured_length}
  \end{spacing}
\end{figure}
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length_error.png}
    \end{center}
    \caption{\textbf{Measured and analytic random walk length relative
        error as a function of the iteration matrix spectral radius.}
      \textit{The weight cutoff was varied with \sn{1}{-2},
        \sn{1}{-4}, and \sn{1}{-8}. The relative error between the
        predicted and measured results is presented for each weight
        cutoff.}}
    \label{fig:measured_length_error}
  \end{spacing}
\end{figure}
